---
title: "Getting Started with socjobmarket"
author: "socjobmarket Package Authors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting Started with socjobmarket}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

The `socjobmarket` package provides a high-performance agent-based simulation framework for analyzing the academic job market in sociology. This vignette will walk you through the basic functionality and show you how to run your first simulations.

## Installation

If you haven't already installed the package:

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("yourusername/socjobmarket")
```

```{r}
# Load the package
library(socjobmarket)
library(ggplot2)  # For additional plotting
```

## Basic Concepts

The simulation models three types of agents:

1. **Graduate Students**: PhD recipients seeking faculty positions
2. **Faculty**: Current faculty members who may retire or move
3. **Departments**: Hiring institutions with varying prestige and standards

The model simulates the job market dynamics from 2000-2020, with special attention to the effects of the 2008 recession.

## Your First Simulation

Let's start with a simple simulation:

```{r}
# Run a single simulation with 20 years of data
results <- run_simulation_optimized(
  seed = 123,           # For reproducibility
  simulation_years = 20,
  annual_phd_cohort = 300,
  num_departments = 150
)

# Look at the structure
str(results, max.level = 1)
```

The results contain:
- `yearly_stats`: Annual job market statistics
- `candidate_outcomes`: Individual candidate career outcomes

```{r}
# View the first few years of data
head(results$yearly_stats)
```

## Visualizing Results

Create a basic plot to see placement rates over time:

```{r}
plot <- create_single_simulation_plot(results)
print(plot)
```

You can also create diagnostic plots to understand market dynamics:

```{r}
diagnostics <- create_diagnostic_plots(results)
print(diagnostics$supply_vs_demand)
```

## Understanding Scenarios

The package includes predefined scenarios to test different hypotheses about post-2008 changes:

```{r}
# View available scenarios
scenarios <- get_predefined_scenarios()
names(scenarios)
```

```{r}
# Look at one scenario in detail
scenarios$retirement_delays_only
```

## Running Scenario Analysis

For rigorous research, you'll want to run multiple simulations to account for randomness:

```{r eval=FALSE}
# Run a small scenario analysis (use more simulations for research)
analysis <- main_scenario_analysis(
  n_sims = 10,              # Use 100+ for research
  use_parallel = TRUE,      # Use all CPU cores
  output_dir = "results/"   # Where to save files
)
```

This would create several output files:
- CSV files with detailed results
- PDF with comprehensive plots
- Summary statistics

## Custom Scenarios

You can also create your own scenarios:

```{r eval=FALSE}
# Define a custom scenario
extreme_scenario <- list(
  name = "Extreme Recession Effects",
  retirement_delay_factor = 0.2,    # Very delayed retirement
  search_standards_inflation = 2.0,  # Much higher standards
  postdoc_duration_multiplier = 3.0, # Very long postdocs
  search_failure_tolerance = 0.8,    # High failure tolerance
  apply_recession_effects = TRUE
)

# Run simulation with custom scenario
custom_results <- run_simulation_with_scenario(
  seed = 456,
  scenario = extreme_scenario
)

# Analyze transition rates
transition_rates <- calculate_transition_rates_detailed(
  custom_results$results,
  custom_results$candidate_outcomes
)
print(transition_rates)
```

## Creating Agents

You can also create and examine individual agents:

```{r}
# Create a sample graduate student
student <- create_graduate_student(id = 1, cohort_year = 2020)
print(student)
```

```{r}
# Create a department
department <- create_department(id = 1, prestige_rank = 25)
print(department)
```

```{r}
# Calculate match quality between them
quality <- calculate_match_quality(student, department)
cat("Match quality:", quality)
```

## Batch Agent Creation

For larger simulations, use the batch creation function:

```{r}
# Create multiple students from 2020 cohort
students <- create_multiple_agents("student", n = 5, cohort_year = 2020)
print(students)
```

## Performance Tips

The package is optimized for performance:

1. **Use vectorized functions**: `calculate_match_quality_vectorized()` is much faster than the single version
2. **Enable parallel processing**: Set `use_parallel = TRUE` for scenario analysis
3. **Use data.table**: All internal operations use `data.table` for speed

```{r eval=FALSE}
# Benchmark: vectorized vs single calculations
library(microbenchmark)

# Create test data
candidates <- create_multiple_agents("student", 100, cohort_year = 2020)
dept <- create_department(1)

# Compare speeds
microbenchmark(
  vectorized = calculate_match_quality_vectorized(candidates, dept),
  single = sapply(1:nrow(candidates), function(i) {
    calculate_match_quality(candidates[i,], dept)
  }),
  times = 10
)
```

## Research Workflow

A typical research workflow might look like:

1. **Exploratory analysis**: Run single simulations to understand dynamics
2. **Scenario development**: Create scenarios based on research hypotheses
3. **Monte Carlo analysis**: Run many simulations for statistical power
4. **Result analysis**: Use built-in summarization and plotting functions
5. **Publication**: Export CSV data and PDF plots for papers

## Next Steps

- See the "Advanced Usage" vignette for more complex analyses
- Check the function documentation for detailed parameter descriptions
- Explore the source code on GitHub for customization ideas
- Use the package for your own research on academic labor markets

## Getting Help

If you encounter issues:

1. Check the function documentation: `?function_name`
2. Look at the GitHub issues page
3. Create a new issue with a reproducible example

Happy simulating!